---
title: "Report Draft : Current approaches for web interactive plots in R"
runtime: shiny
output: html_document
---

```{r, echo = FALSE, message = FALSE, warning=FALSE}
#load the following packages:
library(plotly, quietly = TRUE)
library(ggvis, quietly = TRUE)
library(crosstalk, quietly = TRUE)
library(shiny, quietly = TRUE)
library(grid, quietly = TRUE)
library(gridSVG, quietly = TRUE)
library(DT, quietly = TRUE)
library(lattice, quietly = TRUE)

income <- read.csv('nzincome.csv', header = TRUE)
```

### This is still a work in progress...

---

## Executive summary

Approaches for creating interactive data visualisations have gained momentum in the present day. However many of these existing tools are limited in achieving interactivity on larger datasets and cannot handle more complex tasks such as customisation, querying and linking between different views of data. By using more flexible tools, we hope to combat these limitations and potentially create more impactful and fluid interactions for users to help them explore data from their own perspective.

---

##  1 Introduction 
The purpose of this report is to investigate current solutions for creating interactive data visualisations in R that can be accessible through the web. Interactive visualisations help inform and explain our data beyond static plots. We aim to identify key similarities and differences between existing tools and find ways to meet user needs.

### Why web interactive?
Interactive plots allow users to explore the data freely. Though it may be presented in a more visually appealing way, it may help explain a topic to a more general audience. As Murray (2012) suggests, static visualisations can only 'offer precomposed 'views' of data', where as interactive plots can provide us with different perspectives. To share our views and findings more easily, the web provides an ideal platform for communicating and exchanging information. It has become accessible to everyone without worrying about technical issues such as device compatibility and installation. We find web interactive visualisations are commonly used in teaching statistics, education, data journalism and is likely to continue to be demanded for in the future. 

The main motivation for this project is to provide higher-level interactivity to iNZight, a data visualisation software from the University of Auckland. This report aims to compare current approaches, before deciding on which would fit best from a developer's perspective, but also meet the needs of a user.

[ NEEDS MORES REFERENCES TO BACK THE NEED FOR INTERACTIVE GRAPHICS.
- iNZight, Books on interactive graphics - Cook and Swayne...e.t.c]

---

## 2 Simple tools for achieving interactive plots in R

Existing tools for creating web interactive plots in R can generally be classified as a class of R packages known as HTMLwidgets. For tools that do not follow that class, the ggvis and the Shiny package are popular alternatives where interactivity is driven in R. These will be discussed along with the limitations they hold.

#### HTMLWidgets 
An HTMLWidget is an R package that allows users to have access to an existing javascript library through bindings between defined R functions and the JavaScript library (RStudio, 2014). These HTMLwidgets can serve different purposes depending on what the original javascript library does, such as Highcharter(Kunst, 2017) and rbokeh(Hafen, 2014) that generates plots using the HighCharters.js API (Highcharts) and the Bokeh.js API respectively, DataTable(Xie, 2016) that generates interactive tables, and Leafet(Cheng and Xie, 2017) for interactive maps.

The main HTMLwidget package that we have looked at in detail is plotly as it has focused on incorporating interactivity on a wide range of plots and is compatible with R packages Shiny and crosstalk (more details are discussed below). The advantages of using an HTMLwidget is that it is easily shareable in a standalone sense, where we can save the plot for later use. 

##### Plotly
Plotly is a graphing library that uses the Plotly.JS API (Plotly Technologies, 2017) that is built upon D3 (Bostock, 2011). It is powerful in the sense that it can convert plots rendered in ggplot2 into interactive plots. It provides basic interactivity including tooltips, zooming and panning, selection of points, and subsetting groups of data as seen in Figure 1. We can also create and combine plots together using the ```subplot()``` function, allowing users to create facetted plots manually. 

```{r, echo = FALSE, comment = NA, warning = FALSE}
plotly::plot_ly(data = iris, x = ~Sepal.Width, y = ~Sepal.Length, color = ~Species, type = "scatter", mode = "markers")
```

Figure 1: plotly plot of the iris dataset

By considering HTMLwidgets alone, we find that these solutions only provide interactive plots quickly to the user with basic functionalities such as tooltips, zooming and subsetting. We can build upon layers of plot objects but they cannot be pulled apart or modified without replotting. It is hard to customise interactions as the functions that create these plots are well defined unless we know the original JavaScript API well. These plots do not provide more information about the data or can be linked to any other plot, but this can be achieved by combining these widgets with crosstalk.

#### Extending interactivity between HTMLwidgets with crosstalk
Crosstalk (Cheng, 2016) is an add-on package that allows HTMLwidgets to cross-communicate with each other. As Cheng (2016) explains, it is designed to link and co-ordinate different views of the same data. Data is converted into a 'shared' object (via V6), which has a corresponding key for each row observation. When selection occurs, crosstalk communicates what has been selected and the bounded HTMLwidgets will respond accordingly. This is all happens on the browser, where crosstalk acts as a 'messenger' between HTMLwidgets.

```{r, message = FALSE, echo = FALSE, warning = FALSE}
shared_iris <- SharedData$new(iris)
p1 <- plot_ly(shared_iris, x = ~Petal.Length, y = ~Petal.Width, color = ~Species, type = "scatter")
p2 <-  plot_ly(shared_iris, x = ~Sepal.Length, y = ~Sepal.Width, color = ~Species, type="scatter")
p <- subplot(p1, p2)
bscols(
  widths = 12, #need to check on this
  p,
  datatable(shared_iris)
  )
```

Figure 4: Linked brushing between two Plotly plots and a data table

```{r, message = FALSE, echo = FALSE, warning = FALSE}

shared_iris <- SharedData$new(iris)
GGally::ggpairs(iris, aes(color = Species)) %>% ggplotly()

```

Figure 5: A ggplot2 (rendered as Plotly) scatterplot matrix linked together with crosstalk

```{r, message = FALSE, echo = FALSE}
shared_income <- SharedData$new(income)
bscols(
  widths = 6,
  list(filter_checkbox("sex", "Gender", shared_income, ~sex, inline  = TRUE),
       filter_slider("weekly_hrs", "Weekly Hours", shared_income, ~weekly_hrs),
       filter_select("ethnicity", "Ethnicity", shared_income, ~ethnicity)),
  plot_ly(shared_income, x = ~weekly_hrs, y = ~weekly_income, color = ~sex, type = "scatter", mode = "markers")
)
```

Figure 6: Filtering in groups 

However, crosstalk has several limitations. As Cheng (2016) points out, the current interactions that it only supports are linked brushing (Figure 4) and filtering (Figure 6) that can only be done on data in a 'row-observation' format. This means that it cannot be used on aggregate data such as linking a histogram to a scatterplot, as illustrated in Figure 5. When we select over points over the scatterplot matrix, the density curves do not change. Furthermore, it only supports a limited number of HTMLwidgets so far - Plotly, dataTable and Leaflet. This is because the implementation of crosstalk is relatively complex. From a developer's point of view, it requires creating bindings between crosstalk and the HTMLwidget itself and customizing interactions accordingly on how it reacts upon selection and filtering. Despite still being under development, it is promising as other HTMLwidget developers have expressed interest in linking their packages with crosstalk to create more informative visualisations.

#### ggvis
Another common data visualisation package is ggvis (Wickham, Chang 2014). This package utilises the Vega JavaScript library to render its plots and uses Shiny to drive some of its interactions. These plots follow the "Grammar of Graphics" and aim to be an interactive visualisation tool for exploratory analysis. This package has an advantage over htmlwidgets as it expands upon using statistical functions for plotting, such as layer_model_predictions() for drawing trendlines using statistical modelling (see Figure 3). Furthermore, because some of the interactions are driven by Shiny, we can add 'inputs' that look similar to Shiny such as sliders and checkboxes to control and filter the plot, but also have the power to add tooltips as seen in Figure 2.

```{r, message = FALSE, echo = FALSE}
ggvis(iris, ~Sepal.Width, ~Sepal.Length, fill = ~Species) %>%
    layer_points() %>%
    add_tooltip(function(iris) paste("Sepal Width: ", iris$Sepal.Width, "\n",
                                     "Sepal Length: ", iris$Sepal.Length))
```

Figure 2: basic ggvis plot with tooltips

```{r, message = FALSE, echo = FALSE}
ggvis(cocaine, ~weight, ~price, fill = ~state) %>%
  layer_points() %>%
  layer_smooths(stroke:= "red", span = input_slider(0.5, 1, value = 1, label = "Span of loess smoother" )) %>%
  layer_model_predictions(stroke:="blue",
                          model = input_select(c("Loess" = "loess", "Linear Model" = "lm", "RLM" = "MASS::rlm"), label = "Select a model"))

```

Figure 3: Change a trendline with a slider and filters using ggvis alone

[Linking a histogram and scatterplot together?]

However, while we are able to achieve indirect interaction, we are limited to basic interactivity as we are not able to link layers of plot objects together. The user also does not have control over where these inputs such as filters and sliders can be placed on the page. To date, the ggvis package is still under development with more features to come in the near future. With ggvis, we can go further by adding basic user interface options such as filters and sliders to control parts of the plot, but only to a certain extent.

When combining different views of data together, we cannot do this with ggvis and htmlwidgets alone. Fortunately, interactivity can be extended with these packages by coupling it with Shiny.

#### Shiny
Shiny is an R package that build web applications through R (RStudio, 2012). It provides a connection of using R as a server and the browser as a client, such that R outputs are rendered on a web page. This allows users to be able to code in R without the need of learning the other main web technologies HTML, CSS and JavaScript. A Shiny app can be viewed links between 'inputs' (what is being sent to R whenever the end user interacts with different parts of the page) and 'outputs' (what the end user sees on the page and updates whenever an input changes).
There are many different ways to use Shiny to create more interactive data visualisations - we can simply just use Shiny to create interactive plots or extend interactivity in HTMLwidgets and other R packages.

```{r}
# add a even simpler Shiny example - slider controlling redraw entire plot of smoother.

```

Figure 7: A simplistic Shiny app

The main advantage of using Shiny is that it establishes a connection to R to allow for statistical computing to occur, while leaving the browser to drive on-plot interactions. This allows us to be able to link different views of data easily. Furthermore, RStudio has provided ways to be able to host and share these shiny apps via a Shiny server. However, we are still limited in the sense that for every time we launch a Shiny app, we do not have access to R as it runs that session. Additionally, Shiny has a tendency to rerender entire objects whenever an 'input' changes as seen in Figure 7. This may lead to unnecessary computations and may slow down the experience for the user. Despite this, it remains a popular tool for creating interactive visualisations.

##### Interactivity with Shiny alone 
Shiny can provide some interactivity to plots. Figure 8 shows linked brushing between facetted plots and a table. With Shiny, we are able to easily link plots together with other objects. This is done simply by attaching a 'plot_brush' input, and using the brushedPoints() function to return what has been selected to R. As we select on parts of the plot, we see this change occur as the other plot and the table updates and renders what has been selected. Other basic interactions include the addition of clicks (plot_click) and hovers (plot_hover).

```{r, message = FALSE, echo = FALSE}

ui <- basicPage(
  plotOutput("plot", brush = "plot_brush", hover = "plot_hover"),
  verbatimTextOutput("text"),
  tableOutput("income_table")
)

server <- function(input, output) {
  output$plot <- renderPlot({
    ggplot(income, aes(x = weekly_hrs, y= weekly_income, color = sex)) + geom_point() + facet_grid(.~ethnicity)
  })
  
  output$text <- renderText({
    paste0("You've hovering on: ", input$plot_hover$x, ', ' , input$plot_hover$y)
  })

  output$income_table <- renderTable({
    brushedPoints(income, input$plot_brush, xvar = "weekly_hrs", yvar = "weekly_income")
  })
}

shinyApp(ui, server)

```

Figure 8: Facetted ggplot with linked brushing and hovers

However, these basic interactive tools only work on base R plots or plots rendered with ggplot2 and work best on scatter plots. This is because the pixel co-ordinates of the plot are correctly mapped to the data. When we try this on a lattice plot as seen below in Figure 9, this mapping condition fails as the co-ordinates system differs between the data and the plot itself. It is possible to create your own mappings to a plot or image, however it may be complex to develop.

```{r, message = FALSE, echo = FALSE}

shinyApp(
  ui <- basicPage(
  plotOutput("plot", click = "plot_click", brush = "plot_brush"),
  verbatimTextOutput("info")
),

server <- function(input, output) {
  output$plot <- renderPlot({
    x <- income$weekly_hrs
    y <- income$weekly_income
    xyplot(y~x,  main = "Lattice scatterplot of nzincome", ylab ="Weekly income", xlab = "Weekly hrs")
  })

  output$info <- renderText({
    paste0("Weekly_hrs=", input$plot_click$x, "\n Weekly_income=", input$plot_click$y)
  })

}
)

```

Figure 9: A lattice plot that fails to produce correct mapping

With Shiny alone, we can achieve some basic interactivity along with user interface options that are outside of the plot. Despite being limited to plot interactions (clicks, brushes and hovers), we can link these plot interactions to other parts that may give us more information about the data. However, these methods only work for plots that are rendered in base R graphics and ggplot2, and cannot be extended onto other R plots.

##### Extending interactivity with Plotly, ggvis and Shiny

Although Shiny is great at facilitating interactions from outside of a plot, it is limited in facilitating interactions within a plot. It does not have all the capabilities that plotly provides. When we combine the two together, more interaction can be achieved with less effort.

```{r, message = FALSE, echo = FALSE}
ui <- fluidPage(
  fluidRow(
    column(4, plotlyOutput("sourceplot")),
    column(4, plotlyOutput("boxplot")),
    column(4, plotlyOutput("histogram"))
  ),
  tableOutput("table")
)

server <- function(input, output) {

  #scatterplot
  output$sourceplot <- renderPlotly({
    plot_ly(income, x = ~weekly_hrs, y = ~weekly_income, color = ~sex, type = "scatter", mode = "markers", source = "source") %>%
      layout(dragmode = "select")
  })

  #render a box plot based upon the selection from scatterplot
  output$boxplot <- renderPlotly({
    s = event_data('plotly_selected', source = "source")
    selected_data <- income[s[["pointNumber"]], ]
    plot_ly(selected_data, x = ~weekly_hrs, type = "box")
  })

  #render a histogram based upon selection from scatterplot
  output$histogram <- renderPlotly({
    s = event_data('plotly_selected', source = "source")
    selected_data <- income[s[["pointNumber"]], ]
    plot_ly(selected_data, x = ~weekly_hrs, type = "histogram")
  })

  #render table results
  output$table <- renderTable({
    event_data('plotly_selected', source = "source")
  })

}

shinyApp(ui, server)

```

Figure 10: a Shiny app with a Plotly plot with linked brushing

Most HTMLwidgets and ggvis have their own way of incorporating plots into a Shiny application. In Figure 10, we can easily embed plots into Shiny using the ```plotlyOutput()``` function. The plotly package also has its own way of co-ordinating linked brushing and in-plot interactions to other Shiny outputs under a function called ```event_data()```. By combining it with Shiny, we are able to link different plots together and to the data itself that is displayed as a table below.These in-plot interactions are very similar to what Shiny provides for base plots and ggplot2. They work well on scatter plots, but not on other kinds of plots that plotly can provide. These can help generate or change different outputs on the page, but not within themselves. By combining the two together, we get 'on-plot' functionalities from the HTMLwidget, with 'off-plot' driven interactions from Shiny. Similarly, when we can combine ggvis and Shiny together we get similar results as seen in Figure 11. ggvis has its own functions (```ggvisOutput()``` and ```linked_brush()```) that allow for similar interactions to be achieved. 

```{r, message = FALSE, echo = FALSE}
shinyApp(
  
  ui <- basicPage(
  #need to use ggvis output for this
  ggvisOutput("plot"),
  tableOutput("brushed_data")
  #ideally, want to show brushed data through the table
),

server <- function(input, output, session) {

  #create an id for each row of data
  iris <- cbind(iris, id = seq_len(nrow(iris)))

  #creating a linked brush object
  lb <- linked_brush(iris$id, "blue")

  #selecting brushed values:
  selected <- lb$selected

  #making the dataset reactive to the brush
  iris_selected <- reactive({
    if (!any(selected())) return(iris)
    iris[selected(), ]
  })

    iris %>%
      ggvis(~Petal.Length, ~Petal.Width, fill = ~Species) %>%
      layer_points(fill.brush := "blue") %>%
      #takes linked brushing input
      lb$input() %>%
      #adds the selected data points on top of the plot
      add_data(iris_selected) %>%
      bind_shiny("plot")

    output$brushed_data <- renderTable({
      iris_selected()
    })

})

```

Figure 11: an example of linked brushing between ggvis plots


##### A comparison to non-web interactive tools:

The interactions that Shiny achieves are not interactions on the plot itself, but rather an interaction driven outside of the plot that causes it to change. With HTMLwidgets and ggvis, we are unable to easily customize our own interactions into the plot. Furthermore, there are complications on trying to link different plots together. Many of the tools discussed above only provide co-ordinated links to scatter plots. This is a major disadvantage in comparison to older non-web interactive tools such as iPlots and Mondrian. 

iPlots is an R package designed to produce interactive graphs for exploratory data analysis using Java. The main features that are included are querying, highlighting and color brushing. It supports a wide range of different graphs including ```imosaic()``` for mosaic plots, ```ipcp()``` for parallel plots, ```ibox()``` for box plots and ```ihist()``` for histograms. It is easy to modify plots generated as we can easily track what plot objects are added or removed through an object list. 

Mondrian() is similar to iPlots, but is more catered for the end user. It also achieves similar results, with a focus on brushing, linking plots together, querying and visualisation of large data sets.


```{r}
#note that this should run in JGR
# need to include code in for this section

```

Figure 12: Several windows of iPlots that all link together

In Figure 12, we can generate several windows of plots that are seamlessly linked together. When we select on a scatterplot, the selection will appear on any of the other plots including box plots, histograms and bar plots. Furthermore, we can query back the selection to find out the data. The performance and efficiency of these plots is uncompromised when dealing with large datasets.

However, both tools require installation of software. Both iPlots and Mondrian require Java in order to run its interface and produce its plots. Despite being able to handle large queries and seamlessly link many plots together, it does not look visually appealling which may deter users to use something with a more modernistic feel like Shiny. Furthermore, the results generated cannot be easily to shared with others. With these tools, we find that there is a gap between existing tools on the web.

After assessing these tools in detail, we can identify the following problems:
- Interactivity with large datasets and querying
- Expanding interactivity to other kinds of R plots (such as lattice plots)
- Linking different kinds of plots in a standalone sense (not just scatter plots)
- Prevent unnecessary drawing in Shiny
to which we discuss some more tools that can help us solve these.

---

## 3 Stretching limitations with more complex tools

Interactions discussed above can be achieved by R users without the knowledge of HTML, CSS and JavaScript. However, we find that more can be achieved if we have some knowledge of these main web technologies. In this section, we discuss more flexible tools that may help us combat these limitations.

#### gridSVG
gridSVG (Murrell and Potter, 2017) is an R package that allows for the conversion of grid graphics in R into SVG (Scalable Vector Graphics). This is powerful because in most cases (including plotly and ggvis), plots are rendered in SVG which makes it easy to attach interactions to specific elements on the page. The advantage of using gridSVG over others is that there is a clear mapping structure between data and svg elements generated. This is not present in HTMLwidgets and their JavaScript libraries, which makes it hard to identify or trace data back to the elements and vice versa. This also explains why it may be complex to customise interactions on the plot. With gridSVG, we can easily add simple javascript to grid elements in R using ```grid.script()```, or attach an external javascript file to it after the SVG plot has been generated as we can easily identify which elements to attach interactions to.

#### Customising simple plot interactions:
A clear limitation that is present in the existing tools discussed previously is the idea of letting the user customise and make their own interactions on the plot.

[insert figure/html link here]

Figure 13: An example of a customised box plot interaction on an iNZight plot using gridSVG, JavaScript 

One such example is highlighting part of a box plot to show certain values between the median and the lower quartile (Figure 13). While this can be easily achieved with more flexible and lower level tools such as gridSVG and custom JavaScript, it is not with the existing tools such as plotly or ggvis as discussed above.


#### Preventing redraws in Shiny using JavaScript messages and gridSVG:

When we interact with the browser alone, most of these interactions are driven by JavaScript. Shiny provides a way of sending messages from the browser back to R through two different functions: shiny

This gives us a way of preventing rerendering through Shiny when it is not necessary to redraw the entire plot, especially when it comes to plots with many data points. An example of this is the change of a trendline on a plot.

```{r}
#insert trendline challenge part 2
```

Figure 14: A replica of Figure 7, but only the trendline changes

When we compare Figure 13 back with Figure 7...? In Figure 7, the plots are rendered in PNG and as a single image.

This solves the problem of being only limited to base plots and ggplot2, as now we can render grid graphics and achieve the same effect, but also customise our own interactions while maintaining a connection between R and the browser using Shiny. However, the main limitation that we face when we render in SVG is that the DOM cannot handle too many SVG elements at once. This compromises the performance and efficiency, which explains why we are limited to smaller datasets. This problem also exists in ggvis as they use SVG to render their plots.

```{r}
#insert trendline challenge 3.1
```

Figure 15: Plot a smoother and return selected values

Another alternative is to use the DOM package... ?

#### An alternative to Shiny: DOM package

The DOM package (Murrell, 2016) is an R package that allows for DOM requests to be sent from R to a browser. It is similar to Shiny as it establishes a connection between R and the browser, however it requires the user to know about the Document Object Model (DOM) and the core web technologies involved. To compare it to Shiny, we have replicated Figure 7 using DOM.


```{r}
#insert DOM example
```

Figure 16: DOM example of Figure 14 - changing a trendline 

The advantages of using DOM over Shiny as seen in this example is that we have control over the entire page. Also, we can continue to modify elements on the page without having to stop the connection (unless we manage to run into errors). This also allows users to still have access to R while the page is running. In Shiny we are unable to use R in a single session or be able to change it without stopping the application entirely.

However, there are several limitations with this package. 

#### Dealing with large datasets

[ Could include a summary table of interactions ]

---

## Conclusion
There is more that can be achieved in expanding interactive graphics to create better data visualisations for users. Despite having many tools available such as HTMLwidgets and Shiny, we find that they all achieve similar results without allowing the user to further extend or customize unless they have knowledge of the main web techologies involved. Options are even more limited when it comes to dealing with large datasets, querying or linking plots which are handled effortlessly with non-web solutions. There is potential in creating more meaningful web visualisations for the future.

---

## References

Murray, S (2013). Interactive Data Visualisation for the Web. Oreilly Chimera. http://chimera.labs.oreilly.com/books/1230000000345/index.html

Cheng J (2016). crosstalk: Inter-Widget Interactivity for HTML Widgets. R package version 1.0.1.
  https://rstudio.github.io/crosstalk/

Joshua Kunst (). highcharter: A Wrapper for the 'Highcharts' Library. R package version 0.5.0.9999.
  http://jkunst.com/highcharter
  
Yihui Xie (2016). DT: A Wrapper of the JavaScript Library 'DataTables'. R package version 0.2.12.
  http://rstudio.github.io/DT
  
Joe Cheng, Bhaskar Karambelkar and Yihui Xie (2017). leaflet: Create Interactive Web Maps with the
  JavaScript 'Leaflet' Library. R package version 1.1.0.9000. http://rstudio.github.io/leaflet/

Carson Sievert, Chris Parmer, Toby Hocking, Scott Chamberlain, Karthik Ram, Marianne Corvellec and
  Pedro Despouy (). plotly: Create Interactive Web Graphics via 'plotly.js'. R package version
  4.5.6.9000. https://plot.ly/r, https://cpsievert.github.io/plotly_book/,
  https://github.com/ropensci/plotly

Winston Chang and Hadley Wickham (2016). ggvis: Interactive Grammar of Graphics. R package version
  0.4.3. https://CRAN.R-project.org/package=ggvis

Winston Chang, Joe Cheng, JJ Allaire, Yihui Xie and Jonathan McPherson (2017). shiny: Web Application
  Framework for R. R package version 1.0.2. https://CRAN.R-project.org/package=shiny

Paul Murrell (2016). DOM: Interact with Web Browser DOM. R package version 0.5.

Paul Murrell and Simon Potter (2017). gridSVG: Export 'grid' Graphics as SVG. R package version 1.5-1.
  https://CRAN.R-project.org/package=gridSVG


Simon Urbanek and Tobias Wichtrey (2013). iplots: iPlots - interactive graphics for R. R package
  version 1.1-7. https://CRAN.R-project.org/package=iplots
  
Bostock, D3


---

## Extras
A few extra terms explained briefly.

The 3 main web technologies:

Hyper Text Markup Language (HTML): language used to describe content on a webpage. Elements are defined in '< >' known as tags.

Cascading Style Sheets(CSS): language that controls the style, appearance and presentation on a webpage

JavaScript: a programming language that is widely used to drive interactivity on the web and other functionalities

Document Object Model (DOM): a programming interface between HTML and XML.

These definitions are referenced from the Mozilla Development Network (MDN).


