---
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    pandoc_args: [
      "--number-sections",
      "--number-offset=1"
    ]
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
#load libraries:
library(crosstalk)
library(plotly)
library(DT)

#load datasets:
income <- read.csv('~/Desktop/datasets/nzincome.csv', header = TRUE)
```
<br>

# An overview of tools for achieving web interactive plots in R

There are many R packages that create different interactive data visualisations. Existing tools for creating standard web interactive plots in R can generally be classified as a class of R packages known as **htmlwidgets**. For tools that do not follow that class, the **ggvis** and the **shiny** package are popular alternatives where interactivity is driven in R. We assess these tools and identify their strengths and limitations.

## htmlwidgets
An 'htmlwidget' is an R package that allows users to have access to an existing JavaScript library through bindings between defined R functions and the JavaScript library (RStudio, 2014). These packages can serve different purposes depending on what the original javascript library does. These include popular graphing libraries such as  Highcharter(Kunst, 2017) and rbokeh(Hafen, 2014), DataTable(Xie, 2016) that generates interactive tables, and Leaflet(Cheng and Xie, 2017) for interactive maps.

The main package that we have looked at in detail is `plotly` as it has focused on incorporating interactivity on a wide range of plots and is compatible with R packages `shiny` and `crosstalk` (more details are discussed below). The advantages of using an htmlwidget is that it is easily shareable in a standalone sense, where we can save the plot for later use. 

### plotly
Plotly is a graphing library that uses the Plotly.JS API (Plotly Technologies, 2017) built upon D3 (Bostock, 2011). It is powerful in the sense that it can convert plots rendered in ggplot2 into interactive plots. It provides basic interactivity including tooltips, zooming and panning, selection of points, and subsetting groups of data as seen in Figure 1. We can also create and combine plots together using the ```subplot()``` function, allowing users to create facetted plots manually. 

```{r, message = FALSE, warning = FALSE, echo = FALSE}
plotly::plot_ly(data = iris, x = ~Sepal.Width, 
                y = ~Sepal.Length, color = ~Species, 
                type = "scatter", mode = "markers")
```

```{r, eval = FALSE}
plotly::plot_ly(data = iris, x = ~Sepal.Width, 
                y = ~Sepal.Length, color = ~Species, 
                type = "scatter", mode = "markers")
```

**Figure 1: plotly plot of the iris dataset**

The purpose of plotly is to simply provide an easy way of visualising data. With its very simple API in R, we can easily generate a standard plot that can be shared and saved as an HTML web page. However, like many other htmlwidgets, we find that these solutions only provide interactive plots quickly to the user with basic functionalities such as tooltips, zooming and subsetting. We can build upon layers of plot objects but they cannot be pulled apart or modified without replotting. It is hard to customise interactions as the functions that create these plots are well defined unless we know the original JavaScript API (in this case, the plotly.js API and D3) well. These plots do not provide more information about the data or can be linked to any other plot, but this can be achieved by combining these widgets with crosstalk.

#### Extending interactivity between htmlwidgets with crosstalk
Crosstalk (Cheng, 2016) is an add-on package that allows htmlwidgets to cross-communicate with each other. As Cheng (2016) explains, it is designed to link and co-ordinate different views of the same data. Data is converted into a V6 'shared' object, which has a corresponding key for each row observation. When selection occurs, crosstalk communicates what has been selected and the bounded htmlwidgets will respond accordingly. This is all happens on the browser, where crosstalk acts as a 'messenger' between these widgets.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#transform our data into a shared object
shared_iris <- SharedData$new(iris)
#generate plots
p1 <- plot_ly(shared_iris, x = ~Petal.Length, 
              y = ~Petal.Width, color = ~Species, type = "scatter")
p2 <-  plot_ly(shared_iris, x = ~Sepal.Length, 
               y = ~Sepal.Width, color = ~Species, type="scatter")
#layout the plots on the page, along with the data table 
p <- subplot(p1, p2)
bscols(
  widths = 12, 
  p,
  datatable(shared_iris)
  )
```

```{r, eval = FALSE, comment = NA}
#transform our data into a shared object
shared_iris <- SharedData$new(iris)
#generate plots
p1 <- plot_ly(shared_iris, x = ~Petal.Length, 
              y = ~Petal.Width, color = ~Species, type = "scatter")
p2 <-  plot_ly(shared_iris, x = ~Sepal.Length, 
               y = ~Sepal.Width, color = ~Species, type="scatter")
#layout the plots on the page, along with the data table 
p <- subplot(p1, p2)
bscols(
  widths = 12, #need to scale accordingly
  p,
  datatable(shared_iris)
  )
```

*Figure 4: Linked brushing between two plotly plots and a data table*

In Figure 4, we have linked two plots generated by plotly with a table generated by the DT package. When we select over a set of points in one of the plots, the table will respond by filtering all the points that have been selected, while this selection is also highlighted on the other plot. Similarly, if we highlight on the other plot, that selection should change and be updated. This creates a form of multi-directional linking between different views of the iris dataset.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
shared_income <- SharedData$new(income)
bscols(
  widths = 6,
  list(filter_checkbox("sex", "Gender", shared_income, ~sex, inline  = TRUE),
       filter_slider("weekly_hrs", "Weekly Hours", shared_income, ~weekly_hrs),
       filter_select("ethnicity", "Ethnicity", shared_income, ~ethnicity)),
  plot_ly(shared_income, x = ~weekly_hrs, y = ~weekly_income, color = ~sex, type = "scatter", mode = "markers")
)
```

```{r, eval = FALSE, comment = NA}
shared_income <- SharedData$new(income)
bscols(
  widths = 6,
  list(filter_checkbox("sex", "Gender", shared_income, ~sex, inline  = TRUE),
       filter_slider("weekly_hrs", "Weekly Hours", shared_income, ~weekly_hrs),
       filter_select("ethnicity", "Ethnicity", shared_income, ~ethnicity)),
  plot_ly(shared_income, x = ~weekly_hrs, y = ~weekly_income, color = ~sex, type = "scatter", mode = "markers")
)
```

*Figure 5: Additional filtering and selection tabs using crosstalk*

In figure 5, crosstalk can also be used for filtering. We can add specific inputs for filtering parts of our data set using sliders, checkboxes, and dropdown menus to allow more control over how we can subset and query our data.

However, crosstalk has several limitations. As Cheng (2016) points out, the current interactions that it only supports are linked brushing (Figure 4) and filtering (Figure 5) that can only be done on data in a 'row-observation' format. This means that it cannot be used on aggregate data such as linking a histogram to a scatterplot, as illustrated in Figure 6 below. When we select over points over the scatterplot matrix, the density curves do not change as it cannot convert the selection into aggregated data. 

```{r, message = FALSE, eval = FALSE}

```

*Figure 6: Scatterplot matrix*

Furthermore, crosstalk only supports a limited number of HTMLwidgets so far - plotly, DT and Leaflet. This is because the implementation of crosstalk is relatively complex. From a developer's point of view, it requires creating bindings between crosstalk and the HTMLwidget itself and customizing interactions accordingly on how it reacts upon selection and filtering. Despite being under development, it is promising as other htmlwidget developers (notably, Kunst with highcharter and Hafen with rbokeh) have expressed interest in linking their packages with crosstalk to create more informative visualisations.

## ggvis
Another common data visualisation package is ggvis (Wickham, Chang 2014). This package utilises the Vega JavaScript library to render its plots and uses shiny to drive some of its interactions. These plots follow the "Grammar of Graphics" and aim to be an interactive visualisation tool for exploratory analysis. This package has an advantage over htmlwidgets as it expands upon using statistical functions for plotting, such as `layer_model_predictions()` for drawing trend lines using statistical modelling (see Figure 3). Furthermore, because some of the interactions are driven by Shiny, we can add 'inputs' that look similar to Shiny such as sliders and checkboxes to control and filter the plot, but also have the power to add tooltips as seen in Figure 2.

**Note: the following are not interactive. They require a shiny back-end to become interactive. These images are static plots.**

```{r, warning = FALSE, message = FALSE}
ggvis(iris, ~Sepal.Width, ~Sepal.Length, fill = ~Species) %>%
    layer_points() %>%
    add_tooltip(function(iris) paste("Sepal Width: ", iris$Sepal.Width, "\n",
                                     "Sepal Length: ", iris$Sepal.Length))
```

```{r, eval = FALSE}
ggvis(iris, ~Sepal.Width, ~Sepal.Length, fill = ~Species) %>%
    layer_points() %>%
    add_tooltip(function(iris) paste("Sepal Width: ", iris$Sepal.Width, "\n",
                                     "Sepal Length: ", iris$Sepal.Length))
```

*Figure: basic ggvis plot with tooltips*

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggvis(cocaine, ~weight, ~price, fill = ~state) %>%
  layer_points() %>%
  layer_smooths(stroke:= "red", span = input_slider(0.5, 1, value = 1, label = "Span of loess smoother" )) %>%
  layer_model_predictions(stroke:="blue",
                          model = input_select(c("Loess" = "loess", "Linear Model" = "lm", "RLM" = "MASS::rlm"), label = "Select a model"))
```

```{r, eval = FALSE}
ggvis(cocaine, ~weight, ~price, fill = ~state) %>%
  layer_points() %>%
  layer_smooths(stroke:= "red", span = input_slider(0.5, 1, value = 1, label = "Span of loess smoother" )) %>%
  layer_model_predictions(stroke:="blue",
                          model = input_select(c("Loess" = "loess", "Linear Model" = "lm", "RLM" = "MASS::rlm"), label = "Select a model"))
```

*Figure: Change a trendline with a slider and filters using ggvis alone*

However, while we are able to achieve indirect interactions, we are limited to basic interactivity as we are not able to link layers of plot objects together. The user also does not have control over where these inputs such as filters and sliders can be placed on the page. We also cannot save these interactions to a web page like htmlwidgets as they are driven by the shiny framework which requires R. There is an option of saving the plot as a static plot, either as an SVG or PNG format. To date, the ggvis package is still under development with more features to come in the near future. With ggvis, we can go further by adding basic user interface options such as filters and sliders to control parts of the plot, but only to a certain extent.

When combining different views of data together, we cannot do this with ggvis and htmlwidgets alone. Fortunately, interactivity can be extended with these packages by coupling it with shiny.

## shiny
**shiny** (RStudio, 2012) is an R package that builds web applications. It provides a connection of using R as a server and the browser as a client, such that R outputs are rendered on a web page. This allows users to be able to code in R without the need of learning the other main web technologies. A Shiny application can be viewed as links between 'inputs' (what is being sent to R whenever the end user interacts with different parts of the page) and 'outputs' (what the end user sees on the page and updates whenever an input changes).

*Figure: from RStudio, explaining how inputs and outputs work*

To show how this works, we have created a simple shiny application that has a slider that controls the smoothness of the trend line. Whenever the user moves the slider, 

```{r, echo = FALSE, eval = FALSE}
# add a even simpler Shiny example - slider controlling redraw entire plot of smoother.

```

*Figure 7: A simplistic shiny application that has a slider to control the smoothness of the trendline*

These applications can become more complex when more inputs and outputs are added. The main advantage of using shiny is that it establishes a connection to R to allow for statistical computing to occur, while leaving the browser to drive on-plot and off-plot interactions. This allows us to be able to link different views of data easily. Furthermore, RStudio has provided ways to be able to host and share these shiny apps via a Shiny server. However, we are still limited in the sense that for every time we launch a Shiny app, we do not have access to R as it runs that session. Additionally, shiny has a tendency to rerender entire objects whenever an 'input' changes as seen in Figure 7. This may lead to unnecessary computations and may slow down the experience for the user. Despite this, it remains a popular tool for creating interactive visualisations.

There are many different ways to use shiny to create more interactive data visualisations - we can simply just use it to create interactive plots or extend interactivity in HTMLwidgets and other R packages.

### Interactivity with shiny alone 
Shiny can provide some interactivity to plots. Figure 8 shows linked brushing between facetted plots and a table. With Shiny, we are able to easily link plots together with other objects. This is done simply by attaching a 'plot_brush' input, and using the brushedPoints() function to return what has been selected to R. As we select on parts of the plot, we see this change occur as the other plot and the table updates and renders what has been selected. Other basic interactions include the addition of clicks (plot_click) and hovers (plot_hover).

```{r, echo = FALSE, message = FALSE, eval = FALSE}

ui <- basicPage(
  plotOutput("plot", brush = "plot_brush", hover = "plot_hover"),
  verbatimTextOutput("text"),
  tableOutput("income_table")
)

server <- function(input, output) {
  output$plot <- renderPlot({
    ggplot(income, aes(x = weekly_hrs, y= weekly_income, color = sex)) + geom_point() + facet_grid(.~ethnicity)
  })
  
  output$text <- renderText({
    paste0("You've hovering on: ", input$plot_hover$x, ', ' , input$plot_hover$y)
  })

  output$income_table <- renderTable({
    brushedPoints(income, input$plot_brush, xvar = "weekly_hrs", yvar = "weekly_income")
  })
}

shinyApp(ui, server)

```

*Figure 8: Facetted ggplot with linked brushing and hovers*

However, these basic interactive tools only work on base R plots or plots rendered with ggplot2 and work best on scatter plots. This is because the pixel co-ordinates of the plot are correctly mapped to the data. When we try this on a lattice plot as seen below in Figure 9, this mapping condition fails as the co-ordinates system differs between the data and the plot itself. It is possible to create your own mappings to a plot or image, however it may be complex to develop.

```{r, echo = FALSE, message = FALSE, eval = FALSE}

shinyApp(
  ui <- basicPage(
  plotOutput("plot", click = "plot_click", brush = "plot_brush"),
  verbatimTextOutput("info")
),

server <- function(input, output) {
  output$plot <- renderPlot({
    x <- income$weekly_hrs
    y <- income$weekly_income
    xyplot(y~x,  main = "Lattice scatterplot of nzincome", ylab ="Weekly income", xlab = "Weekly hrs")
  })

  output$info <- renderText({
    paste0("Weekly_hrs=", input$plot_click$x, "\n Weekly_income=", input$plot_click$y)
  })

}
)

```

*Figure 9: A lattice plot that fails to produce correct mapping*

With Shiny alone, we can achieve some basic interactivity along with user interface options that are outside of the plot. Despite being limited to plot interactions (clicks, brushes and hovers), we can link these plot interactions to other parts that may give us more information about the data. However, these methods only work for plots that are rendered in base R graphics and ggplot2, and cannot be extended onto other R plots. 
Because the plots displayed are in an image format, we can only view these plots as a single object and cannot pull apart elements on the plot. We are unable to further extend and add onto a plot, such as add a trendline when brushing or change colors of points when clicked on.

### Linking plotly or ggvis with Shiny

Although Shiny is great at facilitating interactions from outside of a plot, it is limited in facilitating interactions within a plot. It does not have all the capabilities that plotly provides. When we combine the two together, more interaction can be achieved with less effort.

```{r, echo = FALSE, message = FALSE, eval = FALSE}
ui <- fluidPage(
  fluidRow(
    column(4, plotlyOutput("sourceplot")),
    column(4, plotlyOutput("boxplot")),
    column(4, plotlyOutput("histogram"))
  ),
  tableOutput("table")
)

server <- function(input, output) {

  #scatterplot
  output$sourceplot <- renderPlotly({
    plot_ly(income, x = ~weekly_hrs, y = ~weekly_income, 
            color = ~sex, type = "scatter", 
            mode = "markers", source = "source") %>%
      layout(dragmode = "select")
  })

  #render a box plot based upon the selection from scatterplot
  output$boxplot <- renderPlotly({
    s = event_data('plotly_selected', source = "source")
    selected_data <- income[s[["pointNumber"]], ]
    plot_ly(selected_data, x = ~weekly_hrs, type = "box")
  })

  #render a histogram based upon selection from scatterplot
  output$histogram <- renderPlotly({
    s = event_data('plotly_selected', source = "source")
    selected_data <- income[s[["pointNumber"]], ]
    plot_ly(selected_data, x = ~weekly_hrs, type = "histogram")
  })

  #render table results
  output$table <- renderTable({
    event_data('plotly_selected', source = "source")
  })

}

shinyApp(ui, server)

```

*Figure 10: a Shiny app with a Plotly plot with linked brushing*

Most HTMLwidgets and ggvis have their own way of incorporating plots into a Shiny application. In Figure 10, we can easily embed plots into Shiny using the ```plotlyOutput()``` function. The plotly package also has its own way of co-ordinating linked brushing and in-plot interactions to other Shiny outputs under a function called ```event_data()```. By combining it with Shiny, we are able to link different plots together and to the data itself that is displayed as a table below.These in-plot interactions are very similar to what Shiny provides for base plots and ggplot2. They work well on scatter plots, but not on other kinds of plots that plotly can provide. These can help generate or change different outputs on the page, but not within themselves. By combining the two together, we get 'on-plot' functionalities from the HTMLwidget, with 'off-plot' driven interactions from Shiny. Similarly, when we can combine ggvis and Shiny together we get similar results as seen in Figure 11. ggvis has its own functions (```ggvisOutput()``` and ```linked_brush()```) that allow for similar interactions to be achieved. 

```{r, echo = FALSE, message = FALSE, eval = FALSE}
shinyApp(
  
  ui <- basicPage(
  #need to use ggvis output for this
  ggvisOutput("plot"),
  tableOutput("brushed_data")
  #ideally, want to show brushed data through the table
),

server <- function(input, output, session) {

  #create an id for each row of data
  iris <- cbind(iris, id = seq_len(nrow(iris)))

  #creating a linked brush object
  lb <- linked_brush(iris$id, "blue")

  #selecting brushed values:
  selected <- lb$selected

  #making the dataset reactive to the brush
  iris_selected <- reactive({
    if (!any(selected())) return(iris)
    iris[selected(), ]
  })

    iris %>%
      ggvis(~Petal.Length, ~Petal.Width, fill = ~Species) %>%
      layer_points(fill.brush := "blue") %>%
      #takes linked brushing input
      lb$input() %>%
      #adds the selected data points on top of the plot
      add_data(iris_selected) %>%
      bind_shiny("plot")

    output$brushed_data <- renderTable({
      iris_selected()
    })

})

```

*Figure 11: an example of linked brushing between ggvis plots*

### A comparison to non-web interactive tools

The interactions that Shiny achieves are not interactions on the plot itself, but rather an interaction driven outside of the plot that causes it to rerender. With HTMLwidgets and ggvis, we are unable to easily customize our own interactions into the plot. Furthermore, there are complications on trying to link different plots together. Many of the tools discussed above only provide co-ordinated links to scatter plots. This is a major disadvantage in comparison to older non-web interactive tools such as iplots and Mondrian. 

iPlots is an R package designed to produce interactive graphs for exploratory data analysis using Java. The main features that are included are querying, highlighting and color brushing. It supports a wide range of different graphs including mosaic plots, parallel plots, box plots and histograms. It is easy to modify plots generated as we can easily track what plot objects are added or removed through an object list. 

```{r, echo = FALSE, eval = FALSE}
#note that this should run in JGR
# need to include code in for this section

```

*Figure 12: Several windows of iPlots that all link together*

In Figure 12, we can generate several windows of plots that are seamlessly linked together. When we select on a scatterplot, the selection will appear on any of the other plots including box plots, histograms and bar plots. Furthermore, we can query back the selection to find out the data. The performance and efficiency of these plots is uncompromised when dealing with large datasets.

However, both tools require installation of software. Both iPlots and Mondrian require Java in order to run its interface and produce its plots. Despite being able to handle large queries and seamlessly link many plots together, it does not look visually appealling which may deter users to use something with a more modernistic feel like Shiny. Furthermore, the results generated cannot be easily to shared with others. With these tools, we find that there is a gap between existing tools on the web.

From assessing these tools, it is difficult to expand on-plot interactions as it requires an in-depth knowledge of the tool itself and how these interactions are defined. We can use these tools for easily visualise our data with standard interactive plots, but if the user wishes to customise interactivity or extend it further, it presents a dead end or a need for learning its respective API. Next, we look at how we could achieve specific on-plot interactions by combining JavaScript with lower levels tools including gridSVG, shiny and DOM.
